---
title: "Analysis"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  size = 'tiny',
  eval = TRUE,                          # TRUE; logical
  echo = TRUE,                          # TRUE; logical
  results = 'markup',                   # 'markup', 'asis', 'hold', 'hide' 
  collapse = TRUE,                      # FALSE; logical
  error = TRUE,                        # TRUE; logical
  message = TRUE,                      # TRUE; logical
  include = TRUE,                       # TRUE; logical
  warning = TRUE,                      # TRUE; logical
  strip.white = TRUE,                   # TRUE; logical
  tidy = TRUE,                          # FALSE; logical
  tidy.opts = list(width.cutoff = 70),  
  prompt = TRUE,                        # FALSE; logical
  comment = NA,                         # '##'; character or NA
  highlight = TRUE,                     # TRUE; logical
  fig.keep = 'high',                    # 'high', 'none', 'all', 'first', 'last'
  fig.show = 'asis',                    # 'asis', 'hold', 'animate', 'hide'
  #fig.width = 6.5,                     # 7; numeric
  #fig.height = 4.5,                    # 7; numeric
  out.width = '0.90\\textwidth',        # character
  out.height = '0.80\\textheight',      # character
  fig.align = 'center',                 # 'default', 'right', 'left', 'center' 
  purl = TRUE,                          # TRUE; logical
  R.options = list(width = 80)
)
```

# Preliminaries

Import libraries to import, manipulate and analyze the dataset

```{r}
library("readr")
options(tidyverse.quiet = TRUE)
library("tidyverse")
## package testthat needed to check whether two R objects are equal/identical
library("testthat", warn.conflicts = FALSE) 
library("multgee", quietly = TRUE)
library("VGAM", quietly = TRUE)
```

Import the dataset and create unique identification variable by combining variables `id` and `vertebra`

```{r}
lesions <- 
  read_csv("xray_spa_dc_synd_Osteo_vert3s.csv", show_col_types = FALSE)
lesions <- 
  lesions |>
  group_by(id, vertebra) |>
  mutate(id_vertebra = cur_group_id()) |>
  ungroup()
```


# Descriptive Statistics
As a sanity check, tried to replicate the descriptive statistics analysis
```{r}
lesions |> 
  _$Osteo_synm_tot_corr_upp_low_2 |> 
  table(useNA = "always")

lesions |> 
  _$Osteo_synm_tot_corr_upp_low_2 |> 
  table() |> 
  prop.table()

lesions |> 
  filter(synm_tot_lag == 1) |> 
  _$Osteo_synm_tot_corr_upp_low_2 |> 
  table(useNA = "always")

lesions |> 
  filter(synm_tot_lag == 1) |> 
  _$Osteo_synm_tot_corr_upp_low_2 |> 
  table() |> 
  prop.table()

lesions |>
  with(table(synm_tot_lag, Osteo_synm_tot_corr_upp_low_2)) |>
  print()

lesions |>
  filter(!is.na(synm_tot_lag)) |>
  _$id |>
  n_distinct()

## This is different from the reported number of 3696
lesions |> 
  _$id_vertebra |> 
  n_distinct()
```
The only different number is the last one, suggesting that the number of unique values `id_vertebra` in the final dataset are `r id_unique = lesions |> _$id_vertebra |> n_distinct(); id_unique` and not 3696. This discrepancy might have occurred due to the way `id_vertebra` was initially constructed.

Finally, we drop variables from the dataset that are not needed
```{r}
lesions <- 
  lesions |>
  select(-id, -bdmard, -synm_tot, -vertebra)
```


# Analysis using `multgee`

The function `nomLORgee` treats the last category as baseline/reference category. Hence, in the previous code `Osteo_synm_tot_corr_upp_low_2 == 2` was treated as the baseline category and the logits that were created were the following two: i) 0 vs 2 and ii) 1 vs 2. Although this does not affect the inference (i.e. the conclusions are identical), the following code will force `Osteo_synm_tot_corr_upp_low_2 == 0` to be the baseline category and the logits will be: i) 1 vs 0 and ii) 2 vs 0 as in `SAS`: 

```{r  tidy=FALSE}
lesions <-
  lesions |>
  mutate(Osteo_synm_tot_corr_upp_low_2 =
           factor(Osteo_synm_tot_corr_upp_low_2, levels = c("1", "2", "0")),
         Osteo_synm_tot_corr_upp_low_2_lag =
           factor(Osteo_synm_tot_corr_upp_low_2_lag),
         synm_tot_lag =
           factor(synm_tot_lag),
         sexe =
           factor(sexe),
         hla =
           factor(hla),
         tabac_10y =
           factor(tabac_10y),
         profession =
           factor(profession),
         bdmard_lag =
           factor(bdmard_lag))
```

## Univariate Analysis
Fitting all univariable models 

```{r}
fit_null <- nomLORgee(formula = Osteo_synm_tot_corr_upp_low_2 ~ 1,
                      data = lesions,
                      id = id_vertebra,
                      repeated = t_new,
                      LORstr = "time.exch")
  
fit_null |> 
  update(formula = Osteo_synm_tot_corr_upp_low_2 ~ synm_tot_lag) |>
  summary()

fit_null |> 
  update(Osteo_synm_tot_corr_upp_low_2 ~ Osteo_synm_tot_corr_upp_low_2_lag) |>
  summary()

fit_null |>
  update(formula = Osteo_synm_tot_corr_upp_low_2 ~ sexe) |>
  summary()

fit_null |>
  update(formula = Osteo_synm_tot_corr_upp_low_2 ~ age_m0) |>
  summary()

fit_null |>
  update(formula = Osteo_synm_tot_corr_upp_low_2 ~ bmi) |>
  summary()

fit_null |>
  update(formula = Osteo_synm_tot_corr_upp_low_2 ~ hla) |>
  summary()

fit_null |>
  update(formula = Osteo_synm_tot_corr_upp_low_2 ~ tabac_10y) |>
  summary()

fit_null |>
  update(formula = Osteo_synm_tot_corr_upp_low_2 ~ profession) |>
  summary()

fit_null |>
  update(formula = Osteo_synm_tot_corr_upp_low_2 ~ bdmard_lag) |>
  summary()
```

## Multivariable Analysis
Fitting the multivariable model

```{r tidy=FALSE}
model_gee_multinomial <- 
  fit_null |>
  update(formula = Osteo_synm_tot_corr_upp_low_2 ~ synm_tot_lag + 
           Osteo_synm_tot_corr_upp_low_2_lag + sexe + age_m0 + bmi + hla + 
           tabac_10y + profession + bdmard_lag) 
model_gee_multinomial |> summary()
```

and the exponentiated coefficients

```{r}
model_gee_multinomial |> 
  coefficients() |> 
  exp()
```


# Additional Checks 

## `nomLORgee`
We first check the function `nomLORgee`. To test its functionality, we compare the GEE model setting `LORstr == "indepedence"` with the corresponding maximum likelihood (ML) model that treats all clusters (and hence observations as independent). According to the theory that the two models should produce identical estimates for the regression parameters and fitted values.

We fit the GEE model with independence local odds ratios structure
```{r}
model_gee_independence <- 
  model_gee_multinomial |>
  update(LORstr = "independence") 
```
and the ML model
```{r}
model_ml <-
  vglm(formula = Osteo_synm_tot_corr_upp_low_2 ~ synm_tot_lag +
         Osteo_synm_tot_corr_upp_low_2_lag + sexe + age_m0 + bmi + hla +
         tabac_10y + profession + bdmard_lag,
       data = lesions,
       family = multinomial(ref = "0"))
```
Their coefficients are identical
```{r}
model_gee_independence_coef <- 
  model_gee_independence |>
  coefficients() |> 
  matrix(12, 2)
model_ml_coef <-
  model_ml |> 
  coefficients() |> 
  matrix(12, 2, TRUE) 
expect_equal(model_gee_independence_coef, model_ml_coef)
```
and so are their fitted values
```{r}
fitted_values_gee <- 
  model_gee_independence |> 
  fitted.values() |>
  as.numeric()
fitted_values_ml <-
  model_ml |> 
  fitted.values() |>
  as.numeric()
expect_equal(fitted_values_gee, fitted_values_ml)
```

The above confirm that the function `nomLORgee` works as expected.

## Different reference categories
The output of the model `model_gee_independence` uses as baseline/reference category category 0. 

The following code fits a gee model with category 2 as baseline/reference category (as done previously but with `LORstr = "independence"`). 
```{r  tidy=FALSE}
lesions2 <-
  lesions |>
  mutate(Osteo_synm_tot_corr_upp_low_2 =
           factor(Osteo_synm_tot_corr_upp_low_2, levels = c("0", "1", "2")))
model_gee_multinomial2 <- 
  nomLORgee(formula = Osteo_synm_tot_corr_upp_low_2 ~ synm_tot_lag +
              Osteo_synm_tot_corr_upp_low_2_lag + sexe + age_m0 + bmi + hla +
              tabac_10y + profession + bdmard_lag,
            data = lesions2,
            id = id_vertebra,
            repeated = t_new,
            LORstr = "independence") 
model_gee_multinomial2 |> summary()
```

Although the parameter estimates differ the two models above provide the same fitted values
```{r}
fitted_values_gee2 <- 
  model_gee_multinomial2 |>
  fitted.values() |>
  _[, c(2, 3, 1)] |>
  as.numeric()
expect_equal(fitted_values_gee, fitted_values_gee2)
```
Hence having different reference categories is not affecting inference.


## Comparison with `SAS`
Finally, I compared the output of the model `model_gee_independence`. According to the theory, the output from `SAS` and `model_gee_independence` should identical. Indeed they are.
```{r}
model_gee_independence |> summary()
```
